Sustainable Pathways for Irish homes in changing TIMES

### 2021-07-01 manuscript.tex
1-July-2021 Intro done, working on housing stock data.

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{amsmath}
\usepackage{comment}
%\usepackage{cite}

@@

Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section. 
\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.

@@

%and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very

@@

\par In Ireland few models have used energy system optimisation model to explore the residential sector. The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector. Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland, some of which are outlined below.

@@

\par In 2015 Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examing new dwelling, to examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock. The model relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%. The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied. BER database filters were used in this model, the BER filters from this study, were updated and applied in TIM. A sensitivity analysis is performed on the effects of internal temperature and rebound. It was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.

@@

Finnaly, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system. Some of the main models used in the NEMF include:
\begin{itemize}
\item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{PLEXOSExemplar}
\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingTob}
\begin{itemize}
\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
\end{itemize}
\item{BioHEAT is a techno-economic simulation model of bioenergy and heat, used by SEAI to support policy decision in Ireland.}\cite{Durusut2018BioHEAT:Sectors}
\end{itemize}

The listed NEMF models and more are used by SEAI to model the impact of incentives and opportunities created by specific policy instruments. The models are a combination of top down, ..........

@@

Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private invesvestment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing a minimum price for electricity over a 15 year period. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. 

@@

\par Sustainable Energy Authority of Ireland (SEAI) have a key role in the rollout of retrofitting and heat pumps in existing homes. In parallel, new dwellings must achieve a minimum of 20\% renewable energy. The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs. As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than building in 2005, and with CAP2019 plans to upgrade 500,000 existing homes to a Building Energy Rating (BER) ‘B2’ by 2030, highlights the energy efficiency first approach in the residential sector. 

@@

% The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands, to provide GHG targets insights both the energy system and residential sector.

%``How should Ireland set GHG targets in the residential sector?” 
%Choosing the Pathways which Create the Least Burden and Offer the Most Opportunity for Ireland

%FROM TIM Paper
%TIM has been co-developed with the LEAP-Ireland model (Mac Uidhir et al., 2020), which is a bottom-up simulation modelof the Irish energy system which simulates the impact of different policy measures and targets on overall GHG emissions inIreland, with a particularly granular representation of transport and residential heat demand. Underlying data for the relevantsectors are shared between TIM and LEAP using a Data Repository and shared coding convention dictionary to improve the consistency between the models, give more robust analytical insights for policy and to share and exchange expertise between themodelling teams. This also facilitates multi-model approaches to energy systems modelling, which can make use of harmonisedhybrid frameworks coupling simulation and optimisation simultaneously (Rogan et al., 2014)
\subsection{TIMES Energy System Optimisation Model}
\label{Overview}
\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
There is a need for a new ESOM to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
\par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution. Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiencies, lifetimes, availability, environmental characteristics and fixed, variable and operation costs. TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.

@@

%With update technological and macroeconomic data, structured specifically to analyse for sectoral five-year carbon budgets towards 2030 and 2050, it is hoped TIM will provide further and more robust insights than previous Irish TIMES model.

@@

%What does new model do that previous didn’t?
%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
%THe implications of net-zero
%Does new policy context require new modelling approach?
%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
%Demands - 
%Improved more detailed Energy Service Demands

\subsection{Residential}
\label{LitR:RES}
The residential sector was developed in parallel with other sectors in TIM, and also been structured concurrently and similarly to the residential sector in the new Irish LEAP model, to allow more accessible soft-linking. Fig.\ref{fig:RES} shows the schematic overview of the residential sector in TIM, illustrating the flow of energy from fuel to demand.  
First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
 \caption{Residential Sector - Reference Energy System}
 \label{fig:RES}
\end{figure}

The 2018 SEAI Energy Balance supplies the base year fuel data, both the quantity and type of fuel. Where 15 residential fuel types are found in the SEAI Energy Balance, TIM uses 10 residential fuel types. For example, bituminous coal, anthracite + manufactured ovoids and lignite \ brown coal briquettes are combined into one fuel called ``Coal" in TIM.  
There are two types of energy services in TIM - Non-Archetype and Archetype. The six non-archetype energy services, are not dependant on the type of dwelling and the base year quantity is the total national consumption of that energy service, this data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database. The demand for non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.
The SEAI BER Public database, which is Ireland's national Energy Performance Certificate (EPC) register, complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD). The four energy service type are derived from the database - space heating, water heating, pumps \& fans and lighting. The space heating calculation (IS EN 13790) used to estimate the expected energy consumption per dwelling, makes some assumptions. The internal temperature assumption has a large error, therefore TIM has made adjustments so that the total consumption of energy services is calibrated to the base year data. IS EN 13790 assumes all living areas ( living room, sitting room and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C. We know this is not the case, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions in TIM.

@@

The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on IS EN 13790. ArDEM-SQL is used to clean the input data in TIM, by changing internal temperatures. The internal temperature is highly dependent on occupancy behaviour, a thermal comfort  thermal discomfort are higher among the over-65s than for other household \cite{Healy2002FuelIreland} - living room temperatues between 12 and 26...s  and therefore there is are large variations per BER rating.
The International Standard ISO 7730 [6] uses these PMV and PPD indices to predict the thermal sensation of people exposed to moderate thermal environments, as well as to specify acceptable thermal environmental conditions for comfort...
Despite this, it is
worth noting that the most inefficient dwellings i.e. band F, managed to achieve an
average temperature during period 2 that was only 0.5°C lower than the most
efficient dwellings i.e. band C.
C = 19.2, D = 18, E = 18.2 and F= 18.7 older occupants dublin Ho
%Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households. In total almost 30% of fuelpoor households had living room temperatures below 18°C and more than two thirds had living room temperatures below 20°C. Overall nearly half of all non-fuel-poor homes had living room temperatures below 20°C. In households with persons aged over 65 years, 16% had living room temperatures below 18°C and over 50% had temperatures below 20°C. On the other end of the scale over 5% of households occupied by persons over 65 years experienced living room temperatures of ≥24°C.https://arrow.tudublin.ie/cgi/viewcontent.cgi?article=1089&context=scienmas
%The BER for the sample dwellings ranged from C1 to F. Over two thirds of the
sample dwellings had a BER of E or F. The remaining dwellings had a BER of C or
D. In total 4 of the sample dwellings had undergone energy efficiency upgrades in
recent years, including cavity wall insulation, attic insulation where applicable and
heating control upgrades. In the case of two sample dwellings, these upgrade
works were completed between monitoring period 1 and 2, and therefore results
could be compared before and after retrofit.

Non-Archetype - lack of data
The fuel is calibrated to the

While also accounting for more fuels and technological options, the residential sector in TIM is structured with around retrofitting, not vice-versa as previous versions, so that it is more aligned to ongoing climate mitigation investments. . The Non-Archetype Demands are so, because of lack of data 
ENERGY CONSUMPTION CHARACTERISTICS of Houses - 
 %The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock. BERs are required for new houses, rented houses, sold houses and renovated houses.

>>>>

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{amsmath}
\usepackage{comment}
\usepackage{booktabs}
%\usepackage{multirow}

%\usepackage[table]{xcolor}
%\usepackage{cite}

@@

Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.

\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.

@@

ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.

@@

\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector. Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland, some of which are outlined below.

@@

\par In 2015 Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwelling, to examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock. The Dwelling Energy Assessment Procedure (DEAP) model used, relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%. DEAP is the national tool for BER calculation and checking building regulation compliance. The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied. BER database filters were used in this model, the BER filters from this study, were updated and applied in TIM. A sensitivity analysis is performed on the effects of internal temperature and rebound. It was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.

@@

Finally, it is worth mentioning SEAI’s National Energy Modelling Framework (NEMF), which is a suite of modelling tools used for modelling the Irish energy system. Some of the main models used in the NEMF include:
\begin{itemize}
\item {Core Structural Model of the Irish Economy (COSMO)  is a macro-economic model used by the Economic and Social Research Institute (ESRI) to create a baseline forecast of energy consumption by sector} \cite{Bergin2017COSMO:Ireland}
\item{PLEXOS is a power systems modeling tool used for electricity modeling and planning}\cite{PLEXOSExemplar}
\item{Energy-efficiency policy analysis tool informs the energy efficiency estimates} \cite{Scheer2015UnlockingTob}
\begin{itemize}
\item{Consumer Choice Model determines the attractiveness of energy efficiency finance options}\cite{SustainableEnergyAuthorityofIreland2017BehaviouralSector}
\end{itemize}
\item{BioHEAT is a techno-economic simulation model of bioenergy and heat, used by SEAI to support policy decision in Ireland.}\cite{Durusut2018BioHEAT:Sectors}
\end{itemize}

The listed models are some of models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.

The models described in this section included a range of national ESOMs and bottom-up residential models. The models have varied objectives such as exploring consumer choices, PV battery systems location, evaluating health effects of retrofitting and integration of bio-energy and heat. The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the every growing BER database is providing high data granularity input to these models. The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures. 
The bottom-up residential models are typically simulation models looking at specific policy. The complexity of the models have increased significantly in the last 20 years, in Ireland the models starting out in Microsoft Excel and switching to Microsoft SQL. 
There has been few studies on the Irish residential sector using ESOMs, as the residential bottom-up models are preferred when focusing on that sector. The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation or transport. The use of ESOM to explore DH in Ireland, allows for insights which the bottom-up simulation models cannot provide. The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.

@@

Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing a minimum price for electricity over a 15 year period. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports to renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. 

@@

\par Sustainable Energy Authority of Ireland (SEAI) have a key role in the rollout of retrofitting and heat pumps in existing homes. In parallel, new dwellings must achieve a minimum of 20\% renewable energy. The ambient heat used in heat pumps is included as a renewable energy, heat pumps are therefore the preferred heating technology despite the high capital costs. As mentioned in Section \ref{Intro:IreRSD}, new buildings in Ireland produce 70\% less GHG emissions than buildings in 2005, along with CAP2019 plans to retrofit 500,000 existing homes to at least a  ‘B2’ energy rating by 2030, highlights the energy efficiency first approach in the residential sector. 

@@

%``How should Ireland set GHG targets in the residential sector?”

\subsection{TIMES Energy System Optimisation Model}
\label{Overview}
\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model, and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
There is a need for a new ESOM to be built specifically to explore Ireland's new 2030 and 2050 climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
\par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution. Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiencies, lifetimes, availability, environmental characteristics and fixed, variable and operation costs. TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.

@@

@@

\subsection{Residential}
\label{LitR:RES}
The new structure and updated technological and macroeconomic data will provide more robust insights on sectoral five-year carbon budgets towards 2030 and 2050 than previous models. The residential sector was developed in parallel with other sectors in TIM, and concurrently with the new Irish LEAP model, which allows for more accessible soft-linking to provide more robust analytical insights by combining simulation and optimisation, and to exchange expertise between modelling teams. Fig.\ref{fig:RES} illustrates the flow of energy from fuel to demand in the residential sector.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
 \caption{Residential Sector - Reference Energy System}
 \label{fig:RES}
\end{figure}

The BER database provides the base year bottom-up energy consumption data, this is extrapolated to represent the entire stock and cross-checked with the top-down energy consumption from the 2018 SEAI Energy Balance. To compare from the two databases, fuels were were combined and condensed down to 13 fuels in TIM. Table \ref{Table:Fuels} shows an overview of the fuels combined from the different data sources.

\begin{table}[htbp]
  \centering
  \footnotesize
  \caption{Residential Fuels}
   \begin{tabular}{lll}
\hline
\textbf{SEAI Energy Balance}  & \textbf{BER Database}  & \textbf{TIM}                  \\ \hline \hline
Bituminous Coal & House   Coal & Coal                  \\
Anthracite + Manufactured & Anthracite &         \\
Lignite \textbackslash Brown Coal  & Manufactured Smokeless    &                       \\ \hline
Sod Peat & Sod Peat & Peat                                 \\
Briquettes  & Peat   Briquettes   &                       \\ \hline
Kerosene & Heating   Oil   & Kerosene              \\
Gasoil / Diesel /DERV   &       &                       \\
Petroleum Coke            &                &         \\ \hline
Natural Gas    & Mains   Gas          & Natural Gas           \\ \hline
Biomass     & Wood   Chips    & Biomass               \\
  & Wood   Logs   &       \\
  & Wood   Pellets (bulk)  &                       \\
 & Wood   Pellets (in bags ) &                       \\ \hline
Solar Thermal &     & Solar                 \\
Solar Photovoltaic             &         &            \\ \hline
Ambient Heat     &   & Ambient Heat (Space)  \\
   &    & Ambient Heat* (Water) \\ \hline
Electricity        & Electricity  & Electricity           \\ \hline
Heat            & Unknown     & Heat                  \\ \hline
LPG     & Bottled   LPG        & LPG                   \\
  & Bulk LPG       &               \\ \hline
Biodiesel      & Biodiesel  & Biodiesel             \\ \hline
  Bioethanol     & Bioethanol   & Ethanol               \\ \hline
 & Solid Multi Fuel    & Coal                  \\
 &      & Peat                  \\
 &     & Wood                  \\ \hline
\end{tabular}
\label{Table:Fuels}
\end{table}

In TIM fuels such as coal, represent a combination of coal fuels from the two different databases. Kerosene accounts for over 70 \% of oil usage, while Gasoil accounts for most of the remaining share, the fuels have similar properties and there was no evidence to support keeping separate fuels would provide insights to residential decarbonisation. The exception is ambient heat, which could potentially provide insights to residential decarbonisation by seperating it into ambient heat for space heating and ambient heat for water heating. The solid multi-fuel from the BER database represents an open fire or stove which is heated by either coal, peat or wood. Through analysis of the SEAI Energy Balance and BER database, the fuel share assigned for coal, peat and wood is 19.4 \%, 74.4 \% and 6.2 \% respectively. 
There are two types of energy services represented in Table \ref{fig:RES} - Non-Archetype and Archetype. Non-archetype energy services fill in the lack of data on archetype energy services. The six non-archetype energy services, are not dependant on the type of dwelling in the model and the base year quantity is the total national consumption of that energy service, this bottom-up data was retrieved from SEAI Residential Energy Report \cite{SustainableEnergyAuthorityofIreland2018EnergySector} and cross-checked with the POTEnCIA (Policy Oriented Tool for Energy and Climate Change Impact Assessment) database and the SEAI Energy Balance, more details on non-archeype services are found in Section \ref{LitR:Lighting,Appliances}. The four archetype energy services are dependent on  building archetype and data is sourced from the BER database, which is discussed in Section \ref{LitR:StockData}. The internal temperature adjustment for space heating is detailed in Section \ref{LitR:StockData}. The demand for archetype and non-archetype energy services is based on the growth of residential dwellings, which is detailed in section \ref{LitR:Demands}.

First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
%What does new model do that previous didn’t?
%TIM residential improvements - DH, HVO, Solar ( in progress ), better graps the cost and savings of retrofits and heat pumps, which can infrom MACC.Build with 2018 baseyear, not 2010 as previous versions had - this suits the 2030 target
%THe implications of net-zero
%Does new policy context require new modelling approach?
%Yes, do account for ongoing investments in climate mitigation opitions, the approach should provide higher granularity for likely pathways
%How does new modelling structure provide insights (give examples) in a way that previous model didn’t?
%Demands - 
%Improved more detailed Energy Service Demands

\subsubsection{House Stock Data}
\label{LitR:StockData}

The BER database, which represents half of the dwelling stock, highlights the affect year of construction has on expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock. BERs are required for new houses, rented houses, sold houses and renovated houses.

ENERGY CONSUMPTION CHARACTERISTICS of Houses - Table?

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which is a reason for Ireland's high excessive winter mortality \cite{Clinch2000HousingMortality}. Fuel Poverty which which is defined as when a \textit{``household spends 10\% or more of income to achieve adequate warmth"} is a major contributor to the high excessive winter deaths. An Element Energy report claimed 28\% of Irish households were in fuel poverty in 2015  \cite{ElementEnergy2015Bottom-upIreland}. Fuel Allowance and the Household Benefits provides support to low income households to alleviate fuel poverty. These payments have been increased over time with a further increase made in Budget 2020, coinciding with the increase in the carbon tax \cite{,EuropeanUnion2020EUObservatory}. Kerr's study finds an explicit link between fuel poverty and social welfare policy in Ireland \cite{,EuropeanUnion2020EUObservatory,Kerr2019PoliticsFrance} and O’Meara highlights the health issues associated with Fuel Poverty \cite{OMeara2016AIreland}. Ironically, it can be argued this places low income households deeply dependent on fossil fuels for health.
\par BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.

The SEAI BER Public database, which is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD). The four energy service type are derived from the database - space heating, water heating, pumps \& fans and lighting. The space heating calculation (IS EN 13790) used to estimate the expected energy consumption per dwelling, makes some assumptions. The internal temperature assumption has a large error, therefore TIM has made adjustments so that the total consumption of energy services is calibrated to the base year data. IS EN 13790 assumes all living areas ( living room, sitting room and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C. We know this is not the case, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions in TIM.

@@

The ArDEM-SQL, which was described in Section \ref{Intro:ESOMIreland} is a simulation model built on IS EN 13790. ArDEM-SQL alters primary energy consumption from the BER database, by changing internal temperatures which are used in TIM. 
%Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households. In total almost 30% of fuelpoor households had living room temperatures below 18°C and more than two thirds had living room temperatures below 20°C. Overall nearly half of all non-fuel-poor homes had living room temperatures below 20°C. In households with persons aged over 65 years, 16% had living room temperatures below 18°C and over 50% had temperatures below 20°C. On the other end of the scale over 5% of households occupied by persons over 65 years experienced living room temperatures of ≥24°C.https://arrow.tudublin.ie/cgi/viewcontent.cgi?article=1089&context=scienmas
%The BER for the sample dwellings ranged from C1 to F. Over two thirds of the

EPC register EU - Ireland
%The quality of Swedish EPC data was initially studied by Göransson [41], subsequently by Claesson [42], Mangold et al. [21] and Hårsman et al. [43], and most recently by Pasichnyi et al. [6]. These studies have found a couple of important deficiencies in Swedish EPC data. First, the energy expert, and the firm he or she belongs to, has been shown to generate variations in Swedish EPC data [43], [44]. More specifically, Hårsman et al. [43] found that the energy expert along with the characteristics of the firm caused a variation of ±20% in the assessment of energy consumption.
%Second, through interviews with energy experts, Claesson [42] found that most EPCs contain uncertain estimates of energy use as well as arbitrarily distributed values—thus challenging the notion that Swedish EPC data quality, while allegedly measured, should be superior to the quality of calculated values in other member states.
CIARAs Paper
Every BER displays over 200 energy related variables per dwelling, and over half of all residential dwellings are accounted. This large and detailed BER database does
BER Database, CSO and Calculations and Filtration 
FIGURE - Expected and Actual Consumption per m2
Than Actual 
DEAP - The space heat use is defined in IS EN 13790 as the heat delivered to the heated space by an ideal heating
system to maintain the set-point temperature during a given period of time. The DEAP calculation of space
heat use is done on a monthly basis based on the procedure described in IS EN 13790. A single-zone
calculation is used, using a single average value of mean internal temperature as described in Section 7.In DEAP, the heating season is defined as running from October to May inclusive. The annual space heat use
is the sum of monthly values for these eight months.
The heating hours and required internal temperatures in DEAP are based on the requirements of a typical
household. The schedule is as follows.
• Weekdays: 07.00 to 09.00 and 17.00 to 23.00
• Weekends: 07.00 to 09.00 and 17.00 to 23.00
This standardised schedule represents a total heating period of 56 hours per week.
